Compositional Quantitative Resource Analysis of C Programs

### 2014-06-26 paper/main.tex
some minor updates in the logic section

When specifying a program, it is often necessary to relate certain
invariants in the pre- and postcondition.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before using the
logic to derive triples, so the Coq theorem for the above example
would look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function body $S_f$ with this
induction hypothesis.  In this process it can be the case that we
have to use the induction hypothesis with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

@@

Our first example on~\pref{fig:xmplmax} derives a precise runtime bound
on a program that searches a maximal element in an array.
The metric used simply counts the assignments performed by the program.
Hence, the resource consumption is closely related to the number $A$ of
times the test \lstinline{a[i] > m} is true during the execution.
%
It is clear that $A+1$ is the count of ``maximums so far in the
array \lstinline{a}'':
$$
  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
$$
where we write $\# S$ for the cardinal of the set $S$.  This number is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our logic it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

The non-trivial part of this derivation is finding the loop
invariant for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program.  It can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
rule is admissible in our system.

The second example on~\pref{fig:xmplbs} shows a use case for
logical variables as well as a metric for stack consumption.
We are interested in showing that a binary search function
$\code{bsearch}$ has logarithmic stack consumption.  We use
a logical variable $Z$ in the function specification
$\htriple{(Z = \log_2(h-l)) + Z \cdot F}{}{Z \cdot F}$
to express that the stack required by the function is fully
returned afterwards.

@@

We already gave an intuition of the meaning of judgements
derived in the logic, to make it formal we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as
\begin{align*}
  \forall \state \, \cost \, m \, \cost' .\,&
  (m \le n \land P(\state) \le \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
  &\implies 0 \le \cost'.
\end{align*}
This predicate is step indexed by an integer $n$ that will
be used in the soundness proof for the function call and
loop cases.  The constraint $0 \le \cost'$ imposed by the
definition makes sure that the program does not stop on a
resource error.  However, it does not rule out memory safety
errors or assertion failures, this is because our logic does
not prove any safety or correctness theorems but only
focuses on resource usage.

The safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcome of a program
statement.  We denote it by $\safeK n B R Q \cont$ and define
it to be
\begin{align*}
& \safe n B {\code{break}} \cont \\
\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
\land& \safe n Q {\code{skip}} \cont.
\end{align*}

We can now define the \emph{semantic validity} of a logic judgement
$B; R \vdash \htriple P S Q$ without function context as follows
\begin{align*}
&\valid n B R P S Q := \\
&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
&\qquad \implies \safe m {P+x} S \cont.
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic, this refinement is necessary to have a stronger
induction hypothesis available during the proof.

Using this semantic validity  we can define the validity of
a function context $\Delta$ that we write $\validC n \Delta$ as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}
%
\noindent
If $S$ is a complete program and $\Delta$ is empty by
expanding the definitions and remarking that $\validC n \Delta$
and $\safeK n B R Q \Kstop$ are true for every $n$ we get that 
$$
  \forall n .\, \safe n P S \Kstop.
$$
This means that for any starting state $\state$, $P(\state)$
provides enough resource for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.

>>>>

When specifying a program, it is often necessary to relate certain
invariants in the pre- and postcondition.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  Because of its flexibility, we used the latter
approach in our development.  This choice makes it possible to have
logical variables almost for free: we can simply use the variable and
binding mechanisms of Coq, our host theory.  When an logical variable
is needed we introduce it using a universal quantifier of Coq before
using the logic to derive triples.  For example, the Coq theorem for
the above example would look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function body $S_f$ with this
induction hypothesis.  In this process it can be the case that we have
to use the induction hypothesis with a different value of a logical
variable (e.g., because the values of the arguments in the recursive
call differ from the values of the arguments of the callee).  To cope
with this problem, assumptions in the function context $\Delta$ are
universally quantified over logical variables.  The {\sc L:Extend}
rule uses the host proof assistant to require that the triple on $f$'s
body is proved for every possible logical variable $y$.

@@

In the following we demonstrate the use of the logic with two example
derivations.

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource consumption is closely related to the
number $A(0)$ of times the test \lstinline{a[i] > m} is true during the
execution.
%
We observe that $A(0)+1$ is the count of ``maximum candidates'' in the
array \lstinline{a} seen by the algorithm. We define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$.  $A(0)$ is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our  it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

The non-trivial part of this derivation is finding the loop
invariant for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximum candidates so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  % As a side remark, if $K \ge 0$, it is possible to
% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
% of the same program.  It can be done either by adapting the proof or by
% applying the {\sc L:Frame} rule one top of the triple already derived.
% More generally, without function calls, the {\sc L:Frame}
% rule is admissible in our system.

The example in~\pref{fig:xmplbs} shows a use case for logical
variables as well as a metric for stack consumption. In a stack
metric, we account a constant cost for a function call ($M_f>0$) that
is returned after the call ($M_r = -M_f <0$).  All other resource cost
are $0$.  We are interested in showing that a binary search function
$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
  + Z \cdot F}{}{Z \cdot F}$ to express that the stack required by the
function is fully returned afterwards.

@@

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as
\begin{align*}
  \forall \state \, \cost \, m \, \cost' .\,&
  (m \le n \land P(\state) \le \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
  &\implies \cost' \ge 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used in the
soundness proof for the function call and loop cases.  The constraint
$\cost' \geq 0$ imposed by the definition ensures that the program does
not stop because of a resource error (recall that a negative resource
counter on the right-hand side is a resource failure).  However, it
does not rule out memory safety errors or assertion failures, this is
because our logic does not prove any safety or correctness theorems
but only focuses on resource usage.

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcome of a program
statement.  We denote it by $\safeK n B R Q \cont$ and define
it to be
\begin{align*}
& \safe n B {\code{break}} \cont \\
\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
\land& \safe n Q {\code{skip}} \cont.
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $\cdot; B; R
\vdash \htriple P S Q$ of the quantitative Hoare logic with
empty function context as follows
\begin{align*}
&\valid n B R P S Q := \\
&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
&\qquad \implies \safe m {P+x} S \cont.
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic, this refinement is necessary to have a stronger
induction hypothesis available during the proof.

Using this semantic validity  we can define the validity of
a function context $\Delta$ that we write $\validC n \Delta$ as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.

XXXXX  Some more explanations.  Where is the difficulty? XXX

%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}

XXXXX  Some more explanations.  Why n+1?XXX

%
\noindent
If $S$ is a complete program and $\Delta$ is empty by
expanding the definitions and remarking that $\validC n \Delta$
and $\safeK n B R Q \Kstop$ are true for every $n$ we get that 
$$
  \forall n .\, \safe n P S \Kstop.
$$
This means that for any starting state $\state$, $P(\state)$
provides enough resource for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.
